The Astropy Project

### 2017-11-18 main.tex
Link directly to astropy project on PyPI

\section*{\textit{Notes and guidelines (to be removed)}}

\begin{itemize}
   \item The goal is to produce a brief and informative paper that covers major
        \astropy principles not mentioned in the first paper, the core v2.0
        package, and infrastructure in \astropy project to support development
        in \python.
    \item We don't plan on including code in this paper, but if you think you
        will need to include code in your section, please add it to a separate
        \python module (.py file) and include it in this repository.
    \item Use \escapecmd{\sectionname} not ``Section,'' \escapecmd{\figurename}
        not ``Figure''
    \item Use \escapecmd{\astropypkg} for the astropy package,
        \escapecmd{\astropy} for the astropy project, \escapecmd{\python} not
        ``Python''
    \item If your subpackage was included in Paper I, then please just include a
        note on what the package does, a reference to paper I, and any new major
        updates to your package
    \item If your subpackage was not included, then please describe the
        subpackage on level with what was in the first paper, and highlight any
        major features in it. Typical length should be equivalent to one page
        (single-column).
    \item Please make sure you are logged into Overleaf or pushing a commit with
        your information to be able to track the contributors to the paper.

\end{itemize}

\section*{Global TODO items}
\begin{itemize}
    \item \sout{subpackage not sub-package!} BMS: checked as done 13/11
    \item spell check!
    \item check consistent astropy usage
\end{itemize}

\section{Introduction} \label{sec:intro}
% first draft by Moritz (hamogu)
% heavily edited by Adrian PW - let me know if you want to revert anything!
All astronomical research makes use of software in some way.  Astronomy as a field has thus long supported the development of software tools
for astronomical tasks: from scripts that enable individual scientific research to software packages for small collaborations to data reduction pipelines for survey operations.
Some software packages are or were supported by large institutions and are intended for a wide range of users. These packages typically provide some level of documentation and user support or
training.
Other packages are developed by individual researchers or research groups and
are then typically used by smaller groups for more domain-specific purposes.
Whether for a package meant for wide distribution or for scripts and programs
for a specific research project, the implementation of astronomical software can
be eased through the use of a library that provides core functionality that is
common to many astronomical tasks.
The users of such software then also benefit from a community and ecosystem
built around a common foundation.
The \astropy project has grown to become this community, and the \astropypkg
core package a feature-rich library for \python-based astronomical software.

\inlinecomment{Tom R}{the 'this community' above is a bit too general - do we mean it is the only community in astronomy, or in the Python astronomy world? I'd change this to either 'to become this community for Python astronomy software' or 'to become such a community'. For instance yt had a community before we came along for the simulation community, so I want to make sure we don't claim we are the only community out there.}
\inlinecomment{Perry G}{Agreed}

The development of the \astropypkg core package began as a largely
community-driven effort to standardize core functionality for astronomical
software in \python.
In this way, its genesis differs from but builds upon many substantial and
% * <bsipocz@gmail.com> 2017-11-13T13:13:08.884Z:
%
% how do astroy builds upon these? or it's more like builds upon the developer experience?
%
%
% ^.
former astronomical software development efforts that were commissioned or
initiated through large institutional support, for example IRAF \citep[developed
at NOAO;][]{IRAF}, MIDAS \citep[developed at ESO;][]{MIDAS}, or Starlink
\citep[originally developed by a consortium of UK institutions and now
maintained by the East Asian Observatory;][]{starlink1982,starlink2013}.
% More recently, community-driven efforts have seen significant success in the astronomical sciences \citep{yt}
\inlinecomment{BMS}{I like the idea to reference yt here}
%
\inlinecomment{Tom R}{Python calls itself a programming language or general-purpose language, not specifically scripting language. I would remove 'scripting' below.}
\inlinecomment{Perry G}{Yes, it's time to drop the scripting adjective; it's far more general than that now.}
\python\footnote{\url{https://www.python.org/}} is an increasingly popular,
general-purpose scripting language that is available under a permissive open
source software license free of charge for all major operating systems.
Stable and well-developed packages provide support for array representation and arithmetic
\citep[\package{numpy};][]{numpy}, a wide variety of functions for scientific
computing \citep[\package{scipy};][]{numpy}, and publication-quality plotting
\citep[\package{matplotlib};][]{matplotlib}. Tens of thousands of other packages
are available which can help with tasks that are not astronomy specific but
might be performed in the course of astronomical research, for example interfacing with databases or statistical inference.

@@

Affiliated packages are listed on the main \astropy website and advertised to the community through \astropy mailing lists, so becoming an affiliated package is a good way for new and existing packages to gain exposure while at the same time promoting \astropy's high standard for code and documentation quality. This process of listing and promoting affiliated packages is one way in which the \astropy project tries to help with \inlinecomment{AG}{Increase?} code reuse in the astronomical community.

@@

%\subsection{Mixin}

@@

The \astropypkg core package is available from several of the \python distributions for scientific
% * <mattwcraig@gmail.com> 2017-10-27T07:13:51.093Z:
%
% > \astropypkg i
% I changed a few of the macros in here to refer to the package rather than the project because it is the package being discussed here.
%
% ^.
computing, such as Anaconda or Enthought Canopy, which include also some of
\astropy's affiliated packages.
Alternatively, \astropypkg can be readily installed from the Python Package
Index\footnote{\url{https://pypi.python.org/pypi}} (PyPI)
\inlinecomment{Tom R}{make footnote link directly to astropy on PyPI?}
using pip, the
recommended command-line tool for installing \python packages
that is included by default since \python 2 $\geq$ 2.7.9 and \python 3 $\geq$ 3.4.

\inlinecomment{Tom R}{is this actually true? I installed Python 3.6 from source recently and it didn't have the pip command readily available. What it does have is the ensurepip module, but we might want to be careful how we phrase this here. Maybe just remove mention of pip being included by default?}
For most recent \astropypkg releases, pre-built wheel distributions are obtainable
from PyPI for Windows, Mac OS X and Linux systems that provide faster
installation compared to the source distribution.
% hamogu: I think that's too detailed for a journal paper.
%     It's in the docs if anyone needs it.
%Optionally, the version of
%the package and the dependencies needed to run or build the software are given
%by a specifier to the pip package manager as described in PEP 440\footnote{For
%more detailed installation instructions and optional flags see
%\url{http://docs.astropy.org/en/stable/install.html\#installing-astropy}}.

@@

\section{Astropy Core Package v2.0}
\label{sec:core}
% Adrian
The \astropy project aims to provide \python-based packages for all tasks that
are commonly needed in a large subset of the astronomical community.
At the foundation is the \astropypkg core package, which provides general
functionality (e.g., coordinate transformations, reading and writing astronomical
files) or base classes for other
packages to utilize for a common interface (e.g., \texttt{NDData}).
In this section, we highlight new features introduced or substantially improved
since version v0.2 (previously described in \citealt{astropy}).

% \subsection{Analytic Functions}
% moved to models and depreciated --
% just here for completeness at the moment
\inlinecomment{BMS}{I suggest to order the subsection lexicographically as reading the draft the current order seemed very random (unlike at the docs page where they the grouping logic is obvious)}
\inlinecomment{Tom R}{the only issue is if we rely on having e.g. units described first if they are used later}

@@

The \texttt{astropy.units} subpackage adds support for representing units and
numbers with associated units --- ``quantities'' --- in code.
Historically, quantities in code have often been represented simply as numbers,
with units implied or noted via comments in the code because of considerations
about speed: having units associated with numbers inherently adds overhead to
numerical operations.
In \texttt{astropy.units}, \texttt{Quantity} objects subclass the \texttt{numpy}
array object and have been designed with speed in mind.

\inlinecomment{Perry G}{Agreeing with Tom below, I would avoid using terms like subclass, decorators, mixin, namespace, (and perhaps even class, instead using objects consistently; this is no place to be pendantic about the distinction between classes and objects; even to the extent of adding a qualifying sentence somewhere like: "In this paper we will use object when we mean object or classes; in software there are important distinctions between the two terms, but they are not relevant in this paper."), and the like and replace them with more generic descriptions, e.g., "extended", or "tools to make defining functions that can use quantities simple to write" }
\inlinecomment{Tom R}{I think saying 'subclass' here is too detailed and is mentioned below in 3.1.1 if needed - 'make it possible to attach units to numpy arrays' or something like that?}

@@

The \texttt{Quantity} object
        is now a subclass of the \texttt{numpy.ndarray} object and therefore
        works well with many of the functions in \texttt{numpy} that support
        array operations. For example, \texttt{Quantity} objects with angular
        units can be passed in to the trigonometric functions implemented in
        \texttt{numpy} (e.g., \texttt{numpy.cos}) and do the right thing: the
        units are internally converted to radians (what the \texttt{numpy}
        trigonometric functions expect) before being passed to \texttt{numpy}.

\subsubsection{Speed improvements}

\inlinecomment{Tom R}{I wonder if we might just want to mention at the top level that performance has been improved in places? This seems a bit too detailed}

The \texttt{Quantity} object is now significantly
        faster than in early versions of \astropypkg. This is in large part due
        to the fact that \texttt{Quantity} subclasses the \texttt{numpy.ndarray}
        object, but even relative to more recent versions of \astropypkg,
        operations with \texttt{Quantity} objects are generally faster. This is
        largely thanks to improvements in the way \texttt{numpy} arrays can be
        subclassed.

@@

\subsubsection{Using quantities in tables}

\inlinecomment{Tom R}{this is a new feature of table, not units, so I think we don't want to mention this here}

        The \texttt{astropy.table} subpackage (see
        \sectionname~\ref{sec:table}) now supports using quantities in tables
        with the \texttt{QTable} class.
        As a result, units specified in FITS tables are supported, and tables
        with quantities can be read and written to various file formats.

@@

The \texttt{astropy.constants} subpackage provides a selection of physical and
astronomical constants as \texttt{Quantity} objects (see
\sectionname~\ref{sec:units}).
A brief description of this package was given in \citep{astropy}.
In version 2.0, the built-in constants have been organized into modules for
specific versions of the constant values.
For example, physical constants have \texttt{codata2014} \citep{codata2014} and
\texttt{codata2010} versions.
Astronomical constants are organized into \texttt{iau2015} and \texttt{iau2012}
modules to indicate their sources (resolutions from the International
Astronomical Union, IAU).
The \texttt{codata2014} and \texttt{iau2015} versions are combined into the
default constant value version: \texttt{astropyconst20}.
For compatibility with \astropypkg version 1.3 and earlier, the \texttt{astropyconst13}
namespace \inlinecomment{Tom R}{jargon alert: 'namespace'} is still available and provides access to the adopted versions of the
constants from earlier versions of \astropypkg.
To use previous versions of the constants as \emph{units} (e.g., solar masses),
the values from the desired namespace have to be imported directly; with version
2.0, \texttt{astropy.units} uses the \texttt{astropyconst20} versions.

@@

\subsubsection{Derivatives of coordinate representations}

\inlinecomment{Tom R}{does this really warrant a separate section or is it just a byproduct of the velocity stuff that can be mentioned in 3.3.2?}

    As mentioned above, the \texttt{Representation} classes act like
    three-dimensional vectors.
    \texttt{astropy.coordinates} now also supports handling first derivatives of
    vectors / representations through the new \texttt{Differential} classes.
    The \texttt{Differential} classes are currently used internally within the
    \texttt{Frame} classes to store the velocity data.

@@

The benchmarks are meant to evolve over time and include an increasing variety of cases. At the moment, the benchmarks are set up as follows - we have generated a standard set of 1000 pairs of random longitudes/latitudes that we use in all benchmarks. Each benchmark is then defined using an input and output coordinate frame, using all combinations of FK4, FK5, Galactic, ICRS and Ecliptic frames. For now we set the epoch of observation to J2000, and the frame to J2000 in the case of the FK5 and ecliptic frames and B1950 for the FK4 frame, but in future we plan to include a larger variety of epochs and equinoxes, as well as tests of conversion to/from Altitude/Azimuth. For each benchmark, we convert the 1000 longitudes/latitudes from the input/output frame with all tools and quantify the comparison by looking at the median, minimum, maximum, and standard deviation of the absolute separation of the output coordinates from each pair of tools. \tablename~\ref{tab:coordinate_benchmarks} gives an example of the relative accuracy of the conversion from FK4 to Galactic coordinates for all pairs of tools. This shows for example that \astropy, Kapteyn and PyTPM agree perfectly (to the precision shown), while PALpy, pySLALIB, and PyAST also agree perfectly amongst themselves, but show an offset of around 0.2\arcsec with the former three packages. Finally, PyEphem disagrees with other packages by 0.4--0.8\arcsec. These values are only meant to be illustrative and will change over time as the benchmarks are refined and packages are updated.
\inlinecomment{Perry G}{Wouldn't the table be better represented as a matrices? Or perhaps as a matrix as a crude image with colors or intensities representing the size of the quantity. Is the difference between median and mean significant enough to warrant showing both? A link to the actual numbers may satisfy readers that want the exact values, but I suspect the majority don't need to see that precision.}
\begin{table}
\begin{center}
\begin{tabular}{llcccc}
\hline
\hline
Package 1 & Package 2 & Median & Mean & Maximum & Std. Dev. \\
          &           & arcsec & arcsec & arcsec & arcsec \\
\hline
astropy  &  kapteyn  & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata  &  palpy    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyast    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyephem  & 0.459  & 0.458 & 0.860 & 0.231 \\
\nodata  &  pyslalib & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pytpm    & 0.000  & 0.000 & 0.000 & 0.000 \\
kapteyn  &  palpy    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyast    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyephem  & 0.459  & 0.458 & 0.860 & 0.231 \\
\nodata  &  pyslalib & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pytpm    & 0.000  & 0.000 & 0.000 & 0.000 \\
palpy    &  pyast    & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata    &  pyephem  & 0.563  & 0.570 & 1.012 & 0.253 \\
\nodata    &  pyslalib & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata    &  pytpm    & 0.218  & 0.196 & 0.249 & 0.056 \\
pyast    &  pyephem  & 0.563  & 0.570 & 1.012 & 0.253 \\
\nodata    &  pyslalib & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata    &  pytpm    & 0.218  & 0.196 & 0.249 & 0.056 \\
pyephem  &  pyslalib & 0.563  & 0.570 & 1.012 & 0.253 \\
\nodata  &  pytpm    & 0.459  & 0.458 & 0.860 & 0.231 \\
pyslalib &  pytpm    & 0.218  & 0.196 & 0.249 & 0.056 \\
\hline
\end{tabular}
\end{center}
\caption{Comparison of the accuracy of the FK4 to Galactic transformation between different packages.\label{tab:coordinate_benchmarks}}
\end{table}

@@

\subsubsection{Barycentric and Heliocentric corrections}
Detailed eclipse or transit
        timing requires accounting for light travel time differences from the
        source to the observatory because of the Earth's motion.
        It is therefore common to instead convert times to the Solar System
        barycenter or heliocenter where the relative timing of photons is
        standardized.
        With the location of a source on the sky (i.e. a \texttt{SkyCoord}
        object), the location of an observatory on Earth (i.e. an
        \texttt{EarthLocation} object), and time values as \texttt{Time}
        objects, the time corrections to shift to the solar system barycenter or
        heliocenter can now be computed with \package{astropy.time} using the
        \texttt{light\_travel\_time} method of a \texttt{Time} object.
\subsubsection{\texttt{Time} objects as \texttt{Table} columns} \texttt{Time} objects
        can now be added as columns in \package{astropy.table} \texttt{Table}
        objects. This is described in more detail in
        \sectionname~\ref{sec:table} with the concept of a ``mixin'' column
        type.

\inlinecomment{Tom R}{I would remove this tiny section. As for units, I think we should only mention the whole mixin stuff in the table section, not for every single object that can go in a table}

@@

\inlinecomment{Perry G}{Some mention that some of these are inspired by Pandas?}
\subsubsection{Support for grouped table operations}
Tables can contain data that naturally form groups, for example monitoring of an object in different bands.
We may then want to split the table or a column into these groups (in this example: observations in different filters)  and
compute some statistic on the data, or filter the groups based on some
criteria.
Grouped operations are now supported by \texttt{Table} objects.
%,
%mainly through the \texttt{group\_by()} method and subsequent operations
%on the grouped version of the table or column.

\subsubsection{Support for table concatenation}
Tables can be combined in several
different ways.
If two tables have the same rows, we may want to stack them
``vertically'' to create a new table with the same columns but all rows.
If two tables are row-matched but have distinct columns, we may want to
stack them ``horizontally'' to create a new table with the same rows
but all columns.
For other situations, more general table concatenations or joins are
also possible when two tables share some common columns.
%All three of these operations are now possible with \texttt{Table}
%objects through the \texttt{vstack}, \texttt{hstack}, and \texttt{join}
%functions in \package{astropy.table}.

\subsubsection{Mixin columns for time and coordinates}
The \texttt{Table} object now supports \texttt{Quantity}, sky coordinate (\texttt{SkyCoord}),
astronomical \texttt{Time} objects and and other objects to serve
as array-like containers for column data through the ``mixin'' protocol.

\inlinecomment{Tom R}{this is waaaaay too short and jargony. It's a very cool feature that deserves more reader-friendly text!}

%hamogu: I've commented this out.
% I think it's very specific and not relevant to non-developers.
% If we want to keep it, we need to add at least 102 senteces about what Pandas
% is and I think that's more text than is warrented for this feature.

%\subsubsection{Transform to and from \texttt{Pandas} \texttt{DataFrame} objects}
%\texttt{Table} objects can now be created from a \texttt{DataFrame}
%object (\texttt{from\_pandas()}), or exported to a \texttt{DataFrame}
%object (\texttt{to\_pandas()}).
%\texttt{DataFrame} objects do not support multi-dimensional array
%columns, so this conversion will fail if the source \texttt{Table}
%object contains any such columns.

@@

%\inlinecomment{SCO}{Pyfits deprecation in favor of io.fits ?}
%\inlinecomment{SCO}{Table and FITS tables}

@@

We summarize key new features or updates to \package{astropy.io} below,
organized by sub-subpackage.

\inlinecomment{Tom R}{is a sub-subpackage a thing, or is it just 'subpackage' all the way down?}

\subsubsection{Enhanced CSV format}

One of the problems when storing a table in an ASCII format is
preserving table meta-data such as comments, keywords and column data
types, units, and descriptions. Using the newly defined \emph{Enhanced
Character Separated Values} (ECSV,  \citet{ape6}) format it is now
possible to write a table to an ASCII-format file and read it back
with no loss of information. The ECSV format has been designed to be
both human-readable and compatible with most simple CSV readers.

\subsubsection{\texttt{.ascii}: fast readers and writers}

The \package{astropy.io.ascii} module now includes a significantly
faster Cython/C engine for reading and writing ASCII files. This is
available for the most common formats.

%following formats: \texttt{basic},
%      \texttt{commented\_header}, \texttt{csv}, \texttt{no\_header},
%      \texttt{rdb}, and \texttt{tab}.
On average the new engine is about
4 to 5 times faster than the corresponding pure-\python
implementation, and is often comparable to the speed of the
\package{Pandas} \citep{pandas} ASCII file interface.  The fast
reader has parallel processing option that allows harnessing
multiple cores for input parsing to achieve even greater speed
gains.

>>>>

\section{Introduction} \label{sec:intro}
% first draft by Moritz (hamogu)
% heavily edited by Adrian PW - let me know if you want to revert anything!
All astronomical research makes use of software in some way.  Astronomy as a field has thus long supported the development of software tools
for astronomical tasks: from scripts that enable individual scientific research to software packages for small collaborations to data reduction pipelines for survey operations.
Some software packages are or were supported by large institutions and are intended for a wide range of users. These packages typically provide some level of documentation and user support or
training.
Other packages are developed by individual researchers or research groups and
are then typically used by smaller groups for more domain-specific purposes.
Whether for a package meant for wide distribution or for scripts and programs
for a specific research project, the implementation of astronomical software can
be eased through the use of a library that provides core functionality that is
common to many astronomical tasks.
The users of such software then also benefit from a community and ecosystem
built around a common foundation.
The \astropy project has grown to become this community for \python astronomy
software, and the \astropypkg core package a feature-rich \python library.

%\inlinecomment{Tom R}{the 'this community' above is a bit too general - do we mean it is the only community in astronomy, or in the Python astronomy world? I'd change this to either 'to become this community for Python astronomy software' or 'to become such a community'. For instance yt had a community before we came along for the simulation community, so I want to make sure we don't claim we are the only community out there.}
%\inlinecomment{Perry G}{Agreed}

The development of the \astropypkg core package began as a largely
community-driven effort to standardize core functionality for astronomical
software in \python.
In this way, its genesis differs from but builds upon many substantial and
% * <bsipocz@gmail.com> 2017-11-13T13:13:08.884Z:
%
% how do astroy builds upon these? or it's more like builds upon the developer experience?
%
%
% ^.
former astronomical software development efforts that were commissioned or
initiated through large institutional support, for example IRAF \citep[developed
at NOAO;][]{IRAF}, MIDAS \citep[developed at ESO;][]{MIDAS}, or Starlink
\citep[originally developed by a consortium of UK institutions and now
maintained by the East Asian Observatory;][]{starlink1982,starlink2013}.
More recently, community-driven efforts have seen significant success in the astronomical sciences \citep{yt}
%\inlinecomment{BMS}{I like the idea to reference yt here}
%
\python\footnote{\url{https://www.python.org/}} is an increasingly popular,
general-purpose programming language that is available under a permissive open
source software license free of charge for all major operating systems.
Stable and well-developed packages provide support for array representation and arithmetic
\citep[\package{numpy};][]{numpy}, a wide variety of functions for scientific
computing \citep[\package{scipy};][]{numpy}, and publication-quality plotting
\citep[\package{matplotlib};][]{matplotlib}. Tens of thousands of other packages
are available which can help with tasks that are not astronomy specific but
might be performed in the course of astronomical research, for example interfacing with databases or statistical inference.

@@

Affiliated packages are listed on the main \astropy website and advertised to the community through \astropy mailing lists, so becoming an affiliated package is a good way for new and existing packages to gain exposure while at the same time promoting \astropy's high standard for code and documentation quality. This process of listing and promoting affiliated packages is one way in which the \astropy project tries to increase code reuse in the astronomical community.

@@

@@

The \astropypkg core package is available from several of the \python distributions for scientific
% * <mattwcraig@gmail.com> 2017-10-27T07:13:51.093Z:
%
% > \astropypkg i
% I changed a few of the macros in here to refer to the package rather than the project because it is the package being discussed here.
%
% ^.
computing, such as Anaconda or Enthought Canopy, which include also some of
\astropy's affiliated packages.
Alternatively, \astropypkg can be readily installed from the Python Package
Index\footnote{\url{https://pypi.org/project/astropy}} (PyPI)
using \texttt{pip}, the standard command-line tool for installing \python packages.
For most recent \astropypkg releases, pre-built wheel distributions are obtainable
from PyPI for Windows, Mac OS X and Linux systems that provide faster
installation compared to the source distribution.
% hamogu: I think that's too detailed for a journal paper.
%     It's in the docs if anyone needs it.
%Optionally, the version of
%the package and the dependencies needed to run or build the software are given
%by a specifier to the pip package manager as described in PEP 440\footnote{For
%more detailed installation instructions and optional flags see
%\url{http://docs.astropy.org/en/stable/install.html\#installing-astropy}}.

@@

\section{Astropy Core Package v2.0}
\label{sec:core}
% Adrian
The \astropy project aims to provide \python-based packages for all tasks that
are commonly needed in a large subset of the astronomical community.
At the foundation is the \astropypkg core package, which provides general
functionality (e.g., coordinate transformations, reading and writing astronomical
files) or base classes for other
packages to utilize for a common interface (e.g., \texttt{NDData}).
In this section, we highlight new features introduced or substantially improved
since version v0.2 (previously described in \citealt{astropy}).  The \astropypkg
provides a full log of changes \footnote{https://github.com/astropy/astropy/blob/master/CHANGES.rst}
over the course of the entire project and more details about individual
subpackages are available in the documentation \footnote{http://docs.astropy.org/en/stable/}.
Beyond what is mentioned below, most subpackages have seen increased performance
since the release of the v0.2 package.

% \subsection{Analytic Functions}
% moved to models and depreciated --
% just here for completeness at the moment
%\inlinecomment{BMS}{I suggest to order the subsection lexicographically as reading the draft the current order seemed very random (unlike at the docs page where they the grouping logic is obvious)}
%\inlinecomment{Tom R}{the only issue is if we rely on having e.g. units described first if they are used later}

@@

The \texttt{astropy.units} subpackage adds support for representing units and
numbers with associated units --- ``quantities'' --- in code.
Historically, quantities in code have often been represented simply as numbers,
with units implied or noted via comments in the code because of considerations
about speed: having units associated with numbers inherently adds overhead to
numerical operations.
In \texttt{astropy.units}, \texttt{Quantity} objects extend \texttt{numpy}
array object and have been designed with speed in mind.

%\inlinecomment{Perry G}{Agreeing with Tom below, I would avoid using terms like subclass, decorators, mixin, namespace, (and perhaps even class, instead using objects consistently; this is no place to be pendantic about the distinction between classes and objects; even to the extent of adding a qualifying sentence somewhere like: "In this paper we will use object when we mean object or classes; in software there are important distinctions between the two terms, but they are not relevant in this paper."), and the like and replace them with more generic descriptions, e.g., "extended", or "tools to make defining functions that can use quantities simple to write" }
%\inlinecomment{Tom R}{I think saying 'subclass' here is too detailed and is mentioned below in 3.1.1 if needed - 'make it possible to attach units to numpy arrays' or something like that?}

@@

The \texttt{Quantity} object extends
the \texttt{numpy.ndarray} object and therefore
works well with many of the functions in \texttt{numpy} that support
array operations. For example, \texttt{Quantity} objects with angular
units can be directly passed in to the trigonometric functions implemented in
\texttt{numpy}. The units are internally converted to radians (what the \texttt{numpy}
trigonometric functions expect) before being passed to \texttt{numpy}.

%\subsubsection{Speed improvements}

%\inlinecomment{Tom R}{I wonder if we might just want to mention at the top level that performance has been improved in places? This seems a bit too detailed}

%The \texttt{Quantity} object is now significantly
        %faster than in early versions of \astropypkg. This is in large part due
        %to the fact that \texttt{Quantity} subclasses the \texttt{numpy.ndarray}
        %object, but even relative to more recent versions of \astropypkg,
        %operations with \texttt{Quantity} objects are generally faster. This is
        %largely thanks to improvements in the way \texttt{numpy} arrays can be
        %subclassed.

@@

@@

The \texttt{astropy.constants} subpackage provides a selection of physical and
astronomical constants as \texttt{Quantity} objects (see
\sectionname~\ref{sec:units}).
A brief description of this package was given in \citep{astropy}.
In version 2.0, the built-in constants have been organized into modules for
specific versions of the constant values.
For example, physical constants have \texttt{codata2014} \citep{codata2014} and
\texttt{codata2010} versions.
Astronomical constants are organized into \texttt{iau2015} and \texttt{iau2012}
modules to indicate their sources (resolutions from the International
Astronomical Union, IAU).
The \texttt{codata2014} and \texttt{iau2015} versions are combined into the
default constant value version: \texttt{astropyconst20}.
For compatibility with \astropypkg version 1.3,  \texttt{astropyconst13}
is available and provides access to the adopted versions of the
constants from earlier versions of \astropypkg.
To use previous versions of the constants as \emph{units} (e.g., solar masses),
the values have to be imported directly; with version
2.0, \texttt{astropy.units} uses the \texttt{astropyconst20} versions.

@@

%\subsubsection{Derivatives of coordinate representations}

%\inlinecomment{Tom R}{does this really warrant a separate section or is it just a byproduct of the velocity stuff that can be mentioned in 3.3.2?}

    %As mentioned above, the \texttt{Representation} classes act like
    %three-dimensional vectors.
    %\texttt{astropy.coordinates} now also supports handling first derivatives of
    %vectors / representations through the new \texttt{Differential} classes.
    %The \texttt{Differential} classes are currently used internally within the
    %\texttt{Frame} classes to store the velocity data.

@@

The benchmarks are meant to evolve over time and include an increasing variety of cases. At the moment, the benchmarks are set up as follows - we have generated a standard set of 1000 pairs of random longitudes/latitudes that we use in all benchmarks. Each benchmark is then defined using an input and output coordinate frame, using all combinations of FK4, FK5, Galactic, ICRS and Ecliptic frames. For now we set the epoch of observation to J2000, and the frame to J2000 in the case of the FK5 and ecliptic frames and B1950 for the FK4 frame, but in future we plan to include a larger variety of epochs and equinoxes, as well as tests of conversion to/from Altitude/Azimuth. For each benchmark, we convert the 1000 longitudes/latitudes from the input/output frame with all tools and quantify the comparison by looking at the median, minimum, maximum, and standard deviation of the absolute separation of the output coordinates from each pair of tools. \tablename~\ref{tab:coordinate_benchmarks} gives an example of the relative accuracy of the conversion from FK4 to Galactic coordinates for all pairs of tools. This shows for example that \astropy, Kapteyn and PyTPM agree perfectly (to the precision shown), while PALpy, pySLALIB, and PyAST also agree perfectly amongst themselves, but show an offset of around 0.2\arcsec with the former three packages. Finally, PyEphem disagrees with other packages by 0.4--0.8\arcsec. These values are only meant to be illustrative and will change over time as the benchmarks are refined and packages are updated.
%\inlinecomment{Perry G}{Wouldn't the table be better represented as a matrices? Or perhaps as a matrix as a crude image with colors or intensities representing the size of the quantity. Is the difference between median and mean significant enough to warrant showing both? A link to the actual numbers may satisfy readers that want the exact values, but I suspect the majority don't need to see that precision.}
\begin{table}
\begin{center}
\begin{tabular}{llcccc}
\hline
\hline
Package 1 & Package 2 & Median & Mean & Maximum & Std. Dev. \\
          &           & arcsec & arcsec & arcsec & arcsec \\
\hline
astropy  &  kapteyn  & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata  &  palpy    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyast    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyephem  & 0.459  & 0.458 & 0.860 & 0.231 \\
\nodata  &  pyslalib & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pytpm    & 0.000  & 0.000 & 0.000 & 0.000 \\
kapteyn  &  palpy    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyast    & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pyephem  & 0.459  & 0.458 & 0.860 & 0.231 \\
\nodata  &  pyslalib & 0.218  & 0.196 & 0.249 & 0.056 \\
\nodata  &  pytpm    & 0.000  & 0.000 & 0.000 & 0.000 \\
palpy    &  pyast    & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata    &  pyephem  & 0.563  & 0.570 & 1.012 & 0.253 \\
\nodata    &  pyslalib & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata    &  pytpm    & 0.218  & 0.196 & 0.249 & 0.056 \\
pyast    &  pyephem  & 0.563  & 0.570 & 1.012 & 0.253 \\
\nodata    &  pyslalib & 0.000  & 0.000 & 0.000 & 0.000 \\
\nodata    &  pytpm    & 0.218  & 0.196 & 0.249 & 0.056 \\
pyephem  &  pyslalib & 0.563  & 0.570 & 1.012 & 0.253 \\
\nodata  &  pytpm    & 0.459  & 0.458 & 0.860 & 0.231 \\
pyslalib &  pytpm    & 0.218  & 0.196 & 0.249 & 0.056 \\
\hline
\end{tabular}
\end{center}
\caption{Comparison of the accuracy of the FK4 to Galactic transformation between different packages.\label{tab:coordinate_benchmarks}}
\end{table}

@@

\subsubsection{Barycentric and Heliocentric corrections}
Detailed eclipse or transit
        timing requires accounting for light travel time differences from the
        source to the observatory because of the Earth's motion.
        It is therefore common to instead convert times to the Solar System
        barycenter or heliocenter where the relative timing of photons is
        standardized.
        With the location of a source on the sky (i.e. a \texttt{SkyCoord}
        object), the location of an observatory on Earth (i.e. an
        \texttt{EarthLocation} object), and time values as \texttt{Time}
        objects, the time corrections to shift to the solar system barycenter or
        heliocenter can now be computed with \package{astropy.time} using the
        \texttt{light\_travel\_time} method of a \texttt{Time} object.

@@

\inlinecomment{Perry G}{Some mention that some of these are inspired by Pandas?}

\subsubsection{Support for grouped table operations}

A table can contain data that naturally forms groups -- for example it may
contain multiple observations of a few sources at different points in time
and in different bands. We may then want to split the table into groups based
on the combination of source observed and the band, and then combine the
results for each combination of source and band in some way (for example finding
the mean or standard deviation of the fluxes or magnitudes over time) or filter
the groups based on user-defined criteria. These kinds of grouping and
aggregation operations are now fully supported by \texttt{Table} objects.

\subsubsection{Support for table concatenation}

\texttt{Table} objects can now be combined in several different ways. If two
tables have the same rows, we may want to stack them ``vertically'' to create a
new table with the same columns but all rows. If two tables are row-matched but
have distinct columns, we may want to stack them ``horizontally'' to create a
new table with the same rows but all columns. For other situations, more general
table concatenations or joins are also possible when two tables share some
common columns.

\subsubsection{Using astropy objects in tables}

The \texttt{Table} object now allows \texttt{Quantity} objects, celestial
coordinate objects (\texttt{SkyCoord}), and date/time objects (\texttt{Time}) to
be used as columns, and also provides a general way for other user-defined
objects to be used as columns. This makes it possible for example to easily
represent catalogs of sources or time series as in Astropy, and having both the
benefits of the \texttt{Table} object (such as accessing specific rows/columns
or groups of rows/columns, combining tables, and so on) and the benefits of e.g.
the \texttt{SkyCoord} or \texttt{Time} classes (such as converting the
coordinates to a different frame, or accessing the date/time as a modified
Julian date or in another time scale).

@@

%\inlinecomment{SCO}{Table and FITS tables}

@@

% We summarize key new features or updates to \package{astropy.io} below,
% organized by sub-subpackage.

% \inlinecomment{Tom R}{is a sub-subpackage a thing, or is it just 'subpackage' all the way down?}

\subsubsection{ASCII}
% \subsubsection{Enhanced CSV format}

One of the problems when storing a table in an ASCII format is
preserving table meta-data such as comments, keywords and column data
types, units, and descriptions. The newly defined \emph{Enhanced
Character Separated Values} (ECSV,  \citet{ape6}) format makes it
possible to write a table to an ASCII-format file and read it back
with no loss of information. The ECSV format has been designed to be
both human-readable and compatible with most simple CSV readers.

% \subsubsection{\texttt{.ascii}: fast readers and writers}

The \package{astropy.io.ascii} subpackage now includes a significantly faster
Cython/C engine for reading and writing ASCII files. This is available for the
most common formats.  On average the new engine is about 4 to 5 times faster
than the corresponding pure-\python implementation, and is often comparable to
the speed of the \package{Pandas} \citep{pandas} ASCII file interface.  The
fast reader has parallel processing option that allows harnessing multiple
cores for input parsing to achieve even greater speed gains.  By default,
\texttt{read()} and \texttt{write()} will attempt to use the fast C engine
when dealing with compatible formats. Certain features of the full read
/ write interface are not available in the fast version, in which case the
pure-\python version will automatically be used.
